Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-16 paper/section-instantiations.tex
Final small changes
New framework setup

We define the model of the indexing theory for the two-dimensional
geometry instantiation as follows. Each of the sorts is interpreted just as
its semantic counterpart: $\semSort{\SynTransl{2}} = \Transl{2}$, $
\semSort{\SynGL{2}} = \GL{2}$, $ \semSort{\SynOrth{2}} = \Orth{2}$,
and $ \semSort{\SynGL{1}} = \GL{1}$.
%  \begin{displaymath}
%    \begin{array}{@{}llll}
%      \semSort{\SynTransl{2}} = \Transl{2}
%      &
%      \semSort{\SynGL{2}} = \GL{2}
%      &
%      \semSort{\SynOrth{2}} = \Orth{2}
%      &
%      \semSort{\SynGL{1}} = \GL{1}
%    \end{array}
%  \end{displaymath}
Each of the index operations (e.g.,~the group structure and
determinant) is interpreted by the intended semantic operation, and
clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. For the
relational interpretations of $\tyPrim{vec}{B,t}$ and
$\tyPrim{real}{s}$, we use the model-based construction from
\autoref{sec:constr-rel-env}, with $R_{\tyPrimNm{vec}}(B,\vec{t}) = \{
(\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \}$ and
$R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}$. The
types of the primitive vector operations presented in
\autoref{sec:motivating-examples} are collected into a context
$\Gamma_\Geom$ with a straightforward interpretation $\eta_\Geom \in
\ctxtSem{\Gamma_\Geom}$.
  % \begin{displaymath}
  %   \begin{array}{l}
  %     R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
  %     R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
  %   \end{array}
  % \end{displaymath}
%   The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
%   %are as expected. % given the informal definitions we gave in
% formalise the informal ones from~\autoref{sec:motivating-examples}, and 
%   the definitions of %We have set
%   $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ 
%   %to relate $\vec{0}$ and $0$ to themselves to 
%   account for the
%   polymorphic zero constants we have assumed.
%\end{example}

\begin{lemma}\label{lem:geom-environments-1}
  For all $\Delta$ and $\rho \in \relEnv{E}_\Geom(\Delta)$, $(\eta_\Geom,\eta_\Geom) \in \rsem{\Gamma_\Geom}{\relEnv{E}_\Geom}\rho$.
\end{lemma}

@@

\subsection{Type Isomorphisms}
\label{sec:types-indexed-abelian-groups}
We say that two types $A$ and $B$ are isomorphic, and write $A \cong
B$, if there exist maps between them that are mutually inverse with
respect to contextual equivalence, i.e., if there are terms $\vdash I
: A \to B$ and $\vdash J : B\to A$ such that $x:A\vdash
J(I(x))\stackrel{ctx}\approx x : A$ and $y:B\vdash
I(J(y))\stackrel{ctx}\approx y : B$.
%\[
%x:A\vdash J(I(x))\stackrel{ctx}\approx x : A
%\text{ and }y:B\vdash I(J(y))\stackrel{ctx}\approx y : B.
%\]
That $\cong$ is a congruence relation on types is straightforward. We
can also derive isomorphisms that are independent of the indexing
theory, such as $A\times B \cong B \times A$, and $\forall
i\mathord:s.A\to B\cong A\to\forall i\mathord:s.B$ for $i$ not free in
$A$.

@@

For proving non-definability results, we consider the general case of
a sort $\mathsf{G}$ with the equational theory of abelian groups. We
assume a group-indexed primitive type $\tyPrimNm{G}$ and set of
primitive operations typed by $\Gamma_\Grp$ that include the group
identity $1:\tyPrim{G}{1}$, operation $(\cdot): \forall
a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
\tyPrim{G}{a\cdot b}$ and inverse $\ ^{-1}: \forall
a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$. The
context $\Gamma_\Grp$ may also include members with types of the form
$\forall \vec{a}\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to ...
\tyPrim{G}{e_n} \to \tyPrim{G}{e}$ as long as the result index $e$ is
generated by the set $\{e_1,...,e_n\}$. One instance is the
sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ on
$\tyPrim{vec}{t}$. For the purposes of non-definability results, we
may choose an arbitrary index-erasure semantics that supports the
primitive operations.

>>>>

We define the model $\mathcal{M}$ of the indexing theory for the two-dimensional
geometry instantiation as follows. Each of the sorts is interpreted just as
its semantic counterpart: $\semSort{\SynTransl{2}}M = \Transl{2}$, $
\semSort{\SynGL{2}}M = \GL{2}$, $ \semSort{\SynOrth{2}}M = \Orth{2}$,
and $ \semSort{\SynGL{1}}M = \GL{1}$.
%  \begin{displaymath}
%    \begin{array}{@{}llll}
%      \semSort{\SynTransl{2}} = \Transl{2}
%      &
%      \semSort{\SynGL{2}} = \GL{2}
%      &
%      \semSort{\SynOrth{2}} = \Orth{2}
%      &
%      \semSort{\SynGL{1}} = \GL{1}
%    \end{array}
%  \end{displaymath}
Each of the index operations (e.g.,~the group structure and
determinant) is interpreted by the intended semantic operation, and
clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. For the
relational interpretations $\mathcal{R}$ of $\tyPrim{vec}{B,t}$ and
$\tyPrim{real}{s}$, we use $\mathcal{R}_{\tyPrimNm{vec}}(B,\vec{t}) = \{
(\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \}$ and
$\mathcal{R}_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}$. The
types of the primitive vector operations presented in
\autoref{sec:motivating-examples} are collected into a context
$\Gamma_\Geom$ with a straightforward interpretation $\eta_\Geom \in
\ctxtSem{\Gamma_\Geom}$.
  % \begin{displaymath}
  %   \begin{array}{l}
  %     R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
  %     R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
  %   \end{array}
  % \end{displaymath}
%   The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
%   %are as expected. % given the informal definitions we gave in
% formalise the informal ones from~\autoref{sec:motivating-examples}, and 
%   the definitions of %We have set
%   $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ 
%   %to relate $\vec{0}$ and $0$ to themselves to 
%   account for the
%   polymorphic zero constants we have assumed.
%\end{example}

\begin{lemma}\label{lem:geom-environments-1}
  For all $\Delta$ and $\rho \in \semSort{\Delta}M$, $(\eta_\Geom,\eta_\Geom) \in \rsem{\Gamma_\Geom}{\relEnv{E}_\Geom}\rho$.
\end{lemma}

@@

\subsection{Type Isomorphisms}
\label{sec:types-indexed-abelian-groups}
We say that two types $A$ and $B$ are isomorphic, and write $A \cong
B$, if there exist maps between them that are mutually inverse with
respect to contextual equivalence, i.e., if there are terms $\vdash I
: A \to B$ and $\vdash J : B\to A$ such that $\ctxteq{\cdot}{x:A}{J(I(x))}{x}{A}$ 
and $\ctxteq{\cdot}{y:B}{I(J(y))} y B$.
%\[
%x:A\vdash J(I(x))\stackrel{ctx}\approx x : A
%\text{ and }y:B\vdash I(J(y))\stackrel{ctx}\approx y : B.
%\]
That $\cong$ is a congruence relation on types is straightforward. We
can also derive isomorphisms that are independent of the indexing
theory, such as $A\times B \cong B \times A$, and $\forall
i\mathord:s.A\to B\cong A\to\forall i\mathord:s.B$ for $i$ not free in
$A$.

@@

For proving non-definability results, we consider the general case of
a sort $\mathsf{G}$ with the equational theory of abelian groups. We
assume a group-indexed primitive type $\tyPrimNm{G}$ and set of
primitive operations typed by $\Gamma_\Grp$ that include the group
identity $1:\tyPrim{G}{1}$, operation $(\cdot): \forall
a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
\tyPrim{G}{a\cdot b}$ and inverse $\ ^{-1}: \forall
a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$. The
context $\Gamma_\Grp$ may also include members with types of the form
$\forall \vec{a}\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to ...
\tyPrim{G}{e_n} \to \tyPrim{G}{e}$ as long as the result index $e$ is
generated by the set $\{e_1,...,e_n\}$. One instance is the
sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ on
$\tyPrim{vec}{t}$. For the purposes of non-definability results, we
may choose an arbitrary index-erasure semantics with $\eta_\Grp \in
\ctxtSem{\Gamma_\Grp}$.
